Topological Equivariant Artist Model for Visualization Library Architecture

### 2022-05-26 paper/main.tex
section renaming + reworking of bundles

\IEEEPARstart{V}isualization software libraries consist of a set of functions for translating data into visual representations. These functions are expected to preserve the structure of the input data, however that structure is specified. Over time, library developers may need to develop new components to adapt to changing data needs, especially the growth of complex structured and distributed data. We propose that the complex structure of data can be described using concepts from algebraic topology and category theory, and that category theory can be used to express the constraints these components must satisfy in order to produce structure preserving data to visual transformations. This abstraction of the data structure and specificaton of the components can then be used to guide the developement of functional\footnote{functional programming paradigm} components. Since these components are functional, they are expected to be more composable and modular than existing components, have fewer side effects and therefore be easier to incorporate alignside existing library elements, and be verifiable and therefore easier to test \cite{huHowFunctionalProgramming2015,hughesWhyFunctionalProgramming1989}. In this paper, we introduce a mathematical framework for describing, constructing, and testing visualization library components and provide a toy example illustrating these concepts.

@@

\subsection{Topology}
\label{sec:related-work:continuity}
\begin{figure}[h!]
  \label{fig:related-work:continuity:ktypes}
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{A topological representation of the data (in purple) is a simplified representation of how the records in a dataset are connected. Discrete rows of a table are discrete points, a 1D continuous function is evaluated on a 1D continuous interval, and a map is 2D continuous surface. }
\end{figure}
The topological properties of a dataset are the ways in which elements are organized, as described by Wilkenson\cite{wilkinsonGrammarGraphics2005}. Some examples, as shown in \autoref{fig:related-work:continuity:ktypes} are discrete rows of a table, a continuous 1D line, or a surface. The topological space (in purple) and the data form a key-value pair, as described by Munzner \cite{munznerVisualizationAnalysisDesign2014}. She imbues semantic meaning to the topology, wherein the key is an independent attribute used as a look up key to retrieve the value in the dataset. She defines the value as a dependent attribute. In this work we use a more general notion of the topological space as an indexing space for the data records. For example the data record in the yellow row in \autoref{fig:related-work:continuity:ktypes} can be expressed as a constant function of the yellow dot; A point on the green line is mapped to by evaluating the gaussian on a point on the interval; and each pixel in the map can be found via a lookup on the 2D surface. This general formulation of topology gives us a language for expressing the continuity of a dataset.

\begin{figure}[h!]
  \label{fig:related-work:continuity:borked}
  \includegraphics[width=1\columnwidth]{continuity.png}
\end{figure}
The continuity matters to visualization libraries because visual algorithms, such as a the line plot and image plot in \autoref{fig:related-work:continuity:borked}, assume that the continuity of their data, as described in taxonomies of visualization algorithms by Troy, M\"{o}ller, and Chi \cite{toryRethinkingVisualizationHighlevel2004,chiTaxonomyVisualizationTechniques2000}. In \autoref{fig:related-work:continuity:borked}, both algorithms are fed the same input matrix. The line plot assumes this matrix is a list of lines and plots it as such. The image map assumes that the data is an image and plots it as such. Neither function has a way of asking the matrix what it is, nor does the matrix encode information about the continuity of the data it contains. While this work is not about choosing the right visualization type for the given continuity, the ability to encode the continuity expectations in the visual algorithm and the data container would provide a way of ensuring that the assumptions match.

@@

\section{Bundles and Sheaves}
\label{sec:atct}
A goal of this work is to develop guidelines for building composable structure preserving functions for translating data into visual representations. Concepts in algebraic topology and category theory provide a language for formally expressing the structure that these visualizations are expected to preserve. Fiber bundles and sheaves, which come from algebraic topolology, provide a uniform generalizable way to describe data and visualization. Category theory is a method to describe how objects are specified, constrained, and composed \cite{wielsManagementEvolvingSpecifications1998}\; therefore in this work we propose categorical constructions of visualization components as specifications for implementable components. A brief introduction to category theory for visualization practitioners is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

\subsection{Fiber Bundles}
\label{sec:atct:fiber-bundles}
Fiber bundles, are a "unified, dimension-independent framework", as described by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, that allow us to seperatly describe the topological properties and fields of a data source and also the mapping between topolology and data values. A fiber bundle $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and continuous surjective map $\pi: \dtotalc \rightarrow \dbasec$\cite{FiberBundle2020}.

@@

@@

\begin{figure}[h!]
  \label{fig:atct:fb}
  \includegraphics[width=\columnwidth]{fb_tau.png}
  \caption{This trivial dataset consists of a single point $\dbasec$ and the fields \texttt{speed, direction, signal, and strength}. Each of these fields has an associated datatype, which is the set of possible valid values for that field. The fiber space is the cartesian product of these sets $\dfiberc$. Each section $\dsectionc$ function returns a single record in the fiber, here the data record outlined in green.  
  \note{maybe change to temperature}}
\end{figure} 
A fiber bundle is a mathematical structure for encoding continuity, type, and datasets with that continuity and type. In \autoref{fig:atct:fb}, the continuity is a discrete point \dbase. The fiber space that encodes the types is the cartesian product of the field types, and the section returns a single record with values of each field type. While this is a very trivial example, a fiber bundle base space can have almost any topology. The fiber space also does not really have restrictions on which types it can encode or the structure of those types - for example, \texttt{speed} and \texttt{direction} are separate fibers in \autoref{fig:atct:fb} but can be combined into a single two dimensional fiber $\dfiber_{speed, direction} = {\mathbb{R}^{+}}^{s}$ The generality in base space and fiber space makes fiber bundles a unified dimension and type independent data abstraction that can encode most of the common visualization data set structures, such as continuous functions, tables, data cubes, images and networks.

@@

\subsection{Input and Output Types}
\label{sec:atct:io}
As described in \autoref{sec:atct:fiber-bundles}, we model data as a map from base space to fiber space.To encapsulate the structure carried in these maps, and provide a way to define these maps in reusable and composable manner, we define categories of base and fiber spaces.

@@

\subsection{Sheaves}
\label{sec:atct:sheaves}
Sheaves on bundles are an "algebraic data structure", as described by Ghrist \cite{ghristElementaryAppliedTopology2014}, for expressing the bookkeeping involved in keeping track of sections of data over opensets. Sheaves are a way of expressing that subsets, distributed data, and streaming data are local sections that can be glued together. Formally, pre-sheaves are functors, meaning they are functions from objects of one category to objects of another category \cite{WhatFunctorDefinitions}. Here the sheaf 
\begin{equation}
  \label{eq:atct:sheaf:functor}
  \sheafc_{\dbasec, \dtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}
\end{equation}
denotes a sheaf from opensets on the base space $\dbase$ to sets of sections on the bundle $\dtotal$. As a contravariant functor it preserves morphism s

@@

\subsubsection{Structure}
\label{sec:atct:sheaves:measurments}
We define structure on data as the allowable \textcolor{action}{transformations on the bundle}, which we formalize as actions $\dfuncc$ on sections of that sheaf. Our construction of $\dfunc$ is a generalization of the classification of measurement scales by their mathematical structure \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}, which often serves as the basis of the equivariance described in \autoref{sec:related-work:equivariance}.

@@

\subsection{Image Functors}
\label{sec:atct:xi}
The data sections defined in \autoref{eq:atct:fb} are sections of sheaves, introduced in \autoref{sec:atct:sheaves}, over open sets $\openset \subseteq \dbase$ in a base space \dbase\ that encodes the topology of the data topology. The graphic sections, defined in \autoref{eq:atct:fb_graphic_section}, are sections of sheaves over open sets $\opensetg \subseteq \gbase$ in a base space \gbase\ that encodes the topology of the graphic. We propose that there exists a continuous \textcolor{functor}{map between topological spaces} \vindexc 
\begin{equation}
  \label{eq:atct:xi}
  \vindexc: \opensetgc \rightarrow \opensetc 
\end{equation}
such that every $\opensetg \subseteq \gbase$ must map to a corresponding set $\openset \subset \dbase$.

@@

\subsection{Defining Maps Between Spaces $\vindex$}
In \autoref{sec:atct:sheaves}, the functors $\vindex, \vindexpull, \vindexpushc$ are introduced to describe the expected relationship between screen and data base spaces. In the construction of the artist, the functor $\vindex$ is defined such that the data base space \dbase\ is a deformation retraction\cite{nlab:deformation_retract} of the graphic space $\gbase$. This means that there is a continuous surjective mapping from every point $\dbasepoint\in\dbase$ to a point $\dbasepoint \in dbase$. To simplify matters, in this paper, we construct the graphic space as a constant multiple of the base space such that 
\begin{equation}
  \underbrace{\dbasec\times[0,1]^{n}}_{\gbasec} \textcolor{functor}{\xmapsto{\hspace{1em}\vindexc\hspace{1em}}} \dbasec
\end{equation}
where n is a thickening of the graphic base space $\gbase$ to account for the dimensionality of the output space
\begin{equation*}
  n = \begin{cases}
    dim(\gbase) - dim(\dbase) & dim(\dbase)<dim(\gbase)\\
  0 & otherwise
  \end{cases}
\end{equation*}
because the data dimensionality $\dbase$ may be too small for a graphic representation.

@@

\subsection{Composite Measurable Components Into Visual Elements}
As shown in \autoref{fig:construction:flow}, the compositor function $\vmarkc$ transforms the measurable components into properties of a visual element. The compositing function $\vmarkc$ transforms the sections of visual elements $\vsectionc$ into sections of graphics $\gsectionc$.
\begin{equation}
  \vmarkc: \cgamma{\dbasec}{\vtotalc} \rightarrow \cgamma{\gbasec}{\gtotalc}
\end{equation}
The compositing function is map from sheaves over $\dbase$ to sheaves over $\gbase$. This is because, as described in \autoref{fig:construction:xi}, the graphic section must be evaluated on all points in the graphic space to generate the visual element corresponding to a data record at a single point $\vartist(\dsection(\dbasepoint)) = \gsection(\vindexpre(\dbasepoint))$.

@@

\subsection{Constructing the Artist: $\vartistc=\vmarkc\circ\vchannelc$}

@@

\section{Discussion: Implementing The Math}
\label{sec:discussion}
The framework specified in \autoref{sec:artist} and \autoref{sec:construction} describes how to build structure preserving visualization components, but it is left to the library developer to follow these guidelines when building and reusing components. 
In this section, we introduce a toy example of building an artist out of the components introduced in \autoref{sec:construction} to illustrate how components that adhere to these specifications are maintainable, extendible, scalable, and support concurrency. Specially, we introduce artists for building different types of bar graphs because it is a visualization type that allows us to demonstrate composability and multivariate data encoding. We build our visualization components by extending the Python visualization library Matplotlib's artist\footnote{Matplotlib artists are our artist's namesake}\cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} to show that components using this model can be incorporated into existing visualization libraries iteratively. While the architecture specified in \autoref{sec:construction} can be implemented fully functionally, we make use of objects to keep track of parameters passed into artists. In this toy example, the small composable components allow for more easily verifying that each component does its transformation correctly before assembling them into larger systems.

\begin{table}[h!]
  \centering
\begin{tabular}{|lcl|}
  \hline \\
   fruit &  calories &  juice \\
  \hline\\
    apple &        95 &   True \\ 
   orange &        67 &   True \\ 
  lemon &        17 &  False \\ 
      lime &        20 &  False \\
  \hline
\end{tabular}
\label{tab:discussion:data}
\end{table}
We construct a toy dataset with a discrete $\dbase$ of 4 points and a fiber space of $\dfiber=\{apple,\, orange,\, lemon\}\times \mathbb{Z}^{+} \times \{\texttt{True}, \texttt{False}\}$. We thinly wrap \autoref{tab:discussion:data} in an object so that the common data interface function is that $\tau = \ttexttt{DataContainerObject.query}$. 
\begin{minted}{python}
  class FruitFrameWrapper:
    def query(self, data_bounds, sampling_rate):
      # local sections are a list of
      # {field: local_batch_of_values}
      return local_sections
\end{minted}

This interface provides a uniform way of accessing subsets of the data, which are local sections. The motivation for a common data interface is that it would allow the artist to talk to different common python data containers, such as numpy\cite{harris2020array}, pandas\cite{jeff_reback_2020_3715232}, xarray \cite{hoyer2017xarray}, and networkx\cite{HagbergExploringNetwork2008}. Currently, data stored in these containers must be unpacked and converted into arrays and matrices in ways that either destroy or recreate the structure encoded in the container. For example a pandas data frame must be unpacked into its columns before it is sent into most artists and continuity is implicit in the columns being the same length rather than a tracked base space $\dbase$.

Because it is more efficient to work with the data in column order, we often treat the data as a collection of single fiber bundles. This is equivalent to the total bundle, as shown in \autoref{eq:artist:operator}. To encode the values in the dataset, we enforce equivariance by writing $\vchannel$ encoders that match the structure of the fields in the dataset. For example, the fruit column is a nominal measurement scale. Therefore we implement a position encoder that respects permutation $\dfunch$ transformations. The most simple form of this $\vchannel$ is a python dictionary that returns an integer position, because Matplotlib's internal parameter space expects a numerical position type. 
\begin{minted}{python}
  def position_encoder(val):
    return {'apple': 0, 'orange': 2, 'lemon': 4, 'lime': 6}[val]
\end{minted}
As mentioned in \autoref{eq:construction:nu:fabrication}, the encoders can be composed up. For example, the compositor $\vchannel$ may need the position to be converted to screen coordinates. Here the screen coordinate $\vchannel$ is a method of a Matplotlib axes object; a Matplotlib axes is akin to a container artist that holds all information about the sub artists plotted within it. 
\begin{minted}{python}
def composite_x_transform(ax, nu):
    return lambda x: ax.transData.transform(
            (position_encoder(x), 0))[0]
\end{minted}
This encoder returns a function that is \texttt{transData.transform} $\vchannel_{transData}$ composed with the position encoder $\vchannel_{position}$ and takes as input a record to be encoded. As with the position encoder, the transData encoder respects permutation transforms because it returns reals; therefore the composite encoder respects permutation transforms. In this model, developers implement $\vchannel$ encoders that are explicit about which $\dfunc_{\vtotal}$ they support. Writing semantically correct encoders is also the responsibility of the developer and is not addressed in the model. For example \mintinline{python}{fruit_encoder = lamda x: {'apple': green, 'orange':'yellow', 'lemon':'red', 'lime':'orange'}} is a valid color encoding with respect to permutation, but none of those colors are intuitive to the data. It is therefore left to the user, or domain specific library developer, to choose $\vchannel$ encoders that are appropriate for their data.

After converting each record into an intermediate visual component $\vsection$, the set of visual records is passed into $\vmarkc$. Here the $\vmarkc$ includes one last encoder, as illustrated in \autoref{eq:construction:q:fabrication}, that assembles the independent visual components into a rectangle. This $\vchannel$ is inside the $\vmarkc$ to hide that library preferred format from the user. It is called \texttt{qhat} to indicate that this is the $\vartist^{\dbase}$ path in \autoref{eq:construction:artist:path}.  This means that the parameters are constructed in data space $\dbase$ and this function returns a pushed forward $\vindexpush\gsection$.

@@

The $\vchannel$ and $\vmark$ are wrapped in a container object that stores the $\vartist = \vmark \circ \vchannel$ composition and a method for computing the $\vsectionc$.
\begin{minted}{python}
  class Bar: 
    def compose_with_nu(self, pfield, ffield, 
          nu, nu_inv:):
        # returns a new copy of the Bar artist 
        # with the additional nu that converts 
        # from a data (F) field value to a 
        # visual (P) field value
        return new

>>>>

\IEEEPARstart{V}isualization software libraries consist of a set of functions for translating data into visual representations. These functions are expected to preserve the structure of the input data, however that structure is specified. Over time, library developers may need to develop new components to adapt to changing data needs, especially the growth of complex structured and distributed data. We propose that the complex structure of data can be described using concepts from algebraic topology and category theory, and that category theory can be used to express the constraints these components must satisfy in order to produce structure preserving data to visual transformations. This abstraction of the data structure and specification of the components can then be used to guide the developement of functional\footnote{functional programming paradigm} components. Since these components are functional, they are expected to be more composable and modular than existing components, have fewer side effects and therefore be easier to incorporate alignside existing library elements, and be verifiable and therefore easier to test \cite{huHowFunctionalProgramming2015,hughesWhyFunctionalProgramming1989}. In this paper, we introduce a mathematical framework for describing, constructing, and testing visualization library components and provide a toy example illustrating these concepts.

@@

\subsection{Topology}
\label{sec:related-work:continuity}
\begin{figure}[h!]
  \label{fig:related-work:continuity:ktypes}
  \includegraphics[width=1\columnwidth]{k_different_types.png}
  \caption{A topological representation of the data (in purple) is a simplified representation of how the records in a dataset are connected. For example, a 1D continuous function is evaluated on a 1D continuous interval, and a map is 2D continuous surface. A network is a union of discrete nodes and continuous lines. The nodes represent the continuity of a table of discrete weather station data and the graph edges are encoded with the distances between the nodes.}
\end{figure}
The topological properties of a dataset are the ways in which elements are organized, as described by Wilkenson\cite{wilkinsonGrammarGraphics2005}. Some examples, as shown in \autoref{fig:related-work:continuity:ktypes} are discrete rows of a table, a continuous 1D line, or a surface. The topological space (in purple) and the data form a key-value pair, as described by Munzner \cite{munznerVisualizationAnalysisDesign2014}. She imbues semantic meaning to the topology, wherein the key is an independent attribute used as a look up key to retrieve the value in the dataset. She defines the value as a dependent attribute. In this work we use a more general notion of the topological space as an indexing space for the data records. For example  in \autoref{fig:related-work:continuity:ktypes}, a point on the green line is mapped to by evaluating the gaussian on a point on the interval and each pixel in the map can be found via a lookup on the 2D surface. This general formulation of topology gives us a language for expressing the continuity of a dataset. The network in \autoref{fig:related-work:continuity:ktypes} is a union of a 0D discrete points and 1D lines connecting those points. The data record in the yellow row can be expressed as a constant function of the yellow network node, while the data on the graph edge between any two points is a function of the distance between those points.

\begin{figure}[h!]
  \label{fig:related-work:continuity:borked}
  \includegraphics[width=1\columnwidth]{continuity.png}
  \caption{The \texttt{line} and \texttt{image} visual algorithms take in the same array of data. The \texttt{line} assumes that the array is a list of lines, while the \texttt{image} assumes that the array is an image.}
\end{figure}
The continuity matters to visualization libraries because visual algorithms, such as a the line plot and image plot in \autoref{fig:related-work:continuity:borked}, assume the continuity of their data, as described in taxonomies of visualization algorithms by Troy, M\"{o}ller, and Chi \cite{toryRethinkingVisualizationHighlevel2004,chiTaxonomyVisualizationTechniques2000}. In \autoref{fig:related-work:continuity:borked}, both \texttt{line} and \textt{image} algorithms are fed the same input matrix. The line plot assumes this matrix is a list of lines and plots it as such. The image map assumes that the data is an image and plots it as such. Neither function has a way of asking the matrix what it is, nor does the matrix encode information about the continuity of the data it contains. While this work is not about choosing the right visualization type for the given continuity, the ability to encode the continuity expectations in the visual algorithm and the data container would provide a way of ensuring that the assumptions match.

@@

\section{Formal Properties of Data \& Graphics}
\label{sec:atct}
A goal of this work is to develop guidelines for building composable structure preserving functions for translating data into visual representations. Concepts in algebraic topology and category theory provide a language for formally expressing the structure that these visualizations are expected to preserve. Fiber bundles and sheaves, which come from algebraic topolology, provide a uniform generalizable way to describe data and visualization. Category theory is a method to describe how objects are specified, constrained, and composed \cite{wielsManagementEvolvingSpecifications1998}\; therefore in this work we propose categorical constructions of visualization components as specifications for implementable components. A brief introduction to category theory for visualization practitioners is in Vickers et al \cite{vickersUnderstandingVisualizationFormal2013}, but they are applying category theory to semantic concerns about visualization design rather than library architecture.

\subsection{Uniform Abstract Data Representation}
\label{sec:atct:fiber-bundles}
Fiber bundles, are a "unified, dimension-independent framework", as described by Butler\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989}, that allow us to seperatly describe the topological properties and fields of a data source and also the mapping between topolology and data values. A fiber bundle $(\dtotalc, \dbasec, \pi, \dfiberc)$ is a structure with topological spaces $\dtotalc, \dfiberc, \dbasec$ and continuous surjective map $\pi: \dtotalc \rightarrow \dbasec$\cite{FiberBundle2020}.

@@

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{fbtau.png}
  \caption{A non-trivial bundle provides a way of expressing data where the fiber changes along the indexing base space. Here $\dfiber_{\dbasepoint_1}$ and $\dfiber_{\dbasepoint_2}$ go in reverse directions \note{do fibers have to be isomorphic?}. The global section $sin$ returns $sin(\dbasepoint_{i})$ which is a value between $[0,1]$
  \label{fig:atct:stripbundle}}
\end{figure}

The bundle in \autoref{fig:atct:stripbundle} is non-trivial, meaning that the fiber is not identical over every point in the base space $\dbase$. One section over this 1D continuous space is a sin function that takes as input a point in the base space $\dbase$ and returns the sin of the point.

@@

\begin{figure}[h!]
  \includegraphics[width=\columnwidth]{fbnet.png}
  \caption{The fibers over the discrete points $\dbase_{1}, \dbase_{2}$  encode the fields in the temperature table and the fiber over the continuous line $\dbase_{3}$ encodes the fields in the distance table \autoref{fig:related-work:continuity:ktypes}. Those tables can be modeled as sections of the bundle in this figure, where the section evaluated at a point in the base space returns the associated row, and on the edge is a constant function that returns the distance.
  \label{fig:atct:networkbundle}}
\end{figure}

The bundle in \autoref{fig:atct:networkbundle} is constructed from a union of bundles over discrete points $\dbase_{1}, \dbase_{3}$ and a 1D continuous connecting line $\dbase_{2}$ that connects the two points.

A fiber bundle is a mathematical structure for encoding continuity, type, and datasets with that continuity and type. In \autoref{fig:atct:fb}, the continuity is a discrete point \dbase. The fiber space that encodes the types is the cartesian product of the field types, and the section returns a single record with values of each field type. While this is a very trivial example, a fiber bundle base space can have almost any topology. The fiber space also does not really have restrictions on which types it can encode or the structure of those types - for example, \texttt{speed} and \texttt{direction} are separate fibers in \autoref{fig:atct:fb} but can be combined into a single two dimensional fiber $\dfiber_{speed, direction} = {\mathbb{R}^{+}}^{s}$ The generality in base space and fiber space makes fiber bundles a unified dimension and type independent data abstraction that can encode most of the common visualization data set structures, such as continuous functions, tables, data cubes, images and networks.

@@

\subsection{Encoding Topology and Field Types}
\label{sec:atct:io}
As described in \autoref{sec:atct:fiber-bundles}, we model data as a map from base space to fiber space.To encapsulate the structure carried in these maps, and provide a way to define these maps in reusable and composable manner, we define categories of base and fiber spaces.

@@

\subsection{Formal Properties of Data Containers}
\label{sec:atct:sheaves}
Sheaves on bundles are an "algebraic data structure", as described by Ghrist \cite{ghristElementaryAppliedTopology2014}, for expressing the bookkeeping involved in keeping track of sections of data over opensets. Sheaves are a way of expressing that subsets, distributed data, and streaming data are local sections that can be glued together. Formally, pre-sheaves are functors, meaning they are functions from objects of one category to objects of another category \cite{WhatFunctorDefinitions}. Here the sheaf 
\begin{equation}
  \label{eq:atct:sheaf:functor}
  \sheafc_{\dbasec, \dtotalc}: \opensetc \rightarrow \cgamma{\opensetc}{\dtotalc\restriction_{\opensetc}}
\end{equation}
denotes a sheaf from opensets on the base space $\dbase$ to sets of sections on the bundle $\dtotal$. As a contravariant functor it preserves morphism s

@@

\subsubsection{Mathematical Structure of Data}
\label{sec:atct:sheaves:measurments}
We define structure on data as the allowable \textcolor{action}{transformations on the bundle}, which we formalize as actions $\dfuncc$ on sections of that sheaf. Our construction of $\dfunc$ is a generalization of the classification of measurement scales by their mathematical structure \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale, thomasMathematizationNotMeasurement2014}, which often serves as the basis of the equivariance described in \autoref{sec:related-work:equivariance}.

@@

\subsection{Associations Between Data and Graphics}
\label{sec:atct:xi}
The data sections defined in \autoref{eq:atct:fb} are sections of sheaves, introduced in \autoref{sec:atct:sheaves}, over open sets $\openset \subseteq \dbase$ in a base space \dbase\ that encodes the topology of the data topology. The graphic sections, defined in \autoref{eq:atct:fb_graphic_section}, are sections of sheaves over open sets $\opensetg \subseteq \gbase$ in a base space \gbase\ that encodes the topology of the graphic. We propose that there exists a continuous \textcolor{functor}{map between topological spaces} \vindexc 
\begin{equation}
  \label{eq:atct:xi}
  \vindexc: \opensetgc \rightarrow \opensetc 
\end{equation}
such that every $\opensetg \subseteq \gbase$ must map to a corresponding set $\openset \subset \dbase$.

@@

\subsection{Map Between Graphics and Data}
In \autoref{sec:atct:sheaves}, the functors $\vindex, \vindexpull, \vindexpushc$ are introduced to describe the expected relationship between screen and data base spaces. In the construction of the artist, the functor $\vindex$ is defined such that the data base space \dbase\ is a deformation retraction\cite{nlab:deformation_retract} of the graphic space $\gbase$. This means that there is a continuous surjective mapping from every point $\dbasepoint\in\dbase$ to a point $\dbasepoint \in dbase$. To simplify matters, in this paper, we construct the graphic space as a constant multiple of the base space such that 
\begin{equation}
  \underbrace{\dbasec\times[0,1]^{n}}_{\gbasec} \textcolor{functor}{\xmapsto{\hspace{1em}\vindexc\hspace{1em}}} \dbasec
\end{equation}
where n is a thickening of the graphic base space $\gbase$ to account for the dimensionality of the output space
\begin{equation*}
  n = \begin{cases}
    dim(\gbase) - dim(\dbase) & dim(\dbase)<dim(\gbase)\\
  0 & otherwise
  \end{cases}
\end{equation*}
because the data dimensionality $\dbase$ may be too small for a graphic representation.

@@

\subsection{Assemble Components Into a Graphic}
As shown in \autoref{fig:construction:flow}, the compositor function $\vmarkc$ transforms the measurable components into properties of a visual element. The compositing function $\vmarkc$ transforms the sections of visual elements $\vsectionc$ into sections of graphics $\gsectionc$.
\begin{equation}
  \vmarkc: \cgamma{\dbasec}{\vtotalc} \rightarrow \cgamma{\gbasec}{\gtotalc}
\end{equation}
The compositing function is map from sheaves over $\dbase$ to sheaves over $\gbase$. This is because, as described in \autoref{fig:construction:xi}, the graphic section must be evaluated on all points in the graphic space to generate the visual element corresponding to a data record at a single point $\vartist(\dsection(\dbasepoint)) = \gsection(\vindexpre(\dbasepoint))$.

@@

\subsection{Implementing the Artist: $\vartistc=\vmarkc\circ\vchannelc$}

@@

\section{Discussion: Translate Spec to Design}
\label{sec:discussion}
The framework specified in \autoref{sec:artist} and \autoref{sec:construction} describes how to build structure preserving visualization components, but it is left to the library developer to follow these guidelines when building and reusing components. In this section, we introduce a toy example of building an artist out of the components introduced in \autoref{sec:construction} to illustrate how components that adhere to these specifications are maintainable, extendible, scalable, and support concurrency.

Specially, we introduce artists for building the graphical elements shown in \autoref{fig:discussion:bar} because it is a visualization type that allows us to demonstrate composability and multivariate data encoding. We build our visualization components by extending the Python visualization library Matplotlib's artist\footnote{Matplotlib artists are our artist's namesake}\cite{hunterMatplotlib2DGraphics2007,hunterArchitectureOpenSource} to show that components using this model can be incorporated into existing visualization libraries iteratively. While the architecture specified in \autoref{sec:construction} can be implemented fully functionally, we make use of objects to keep track of parameters passed into artists. In this toy example, the small composable components allow for more easily verifying that each component does its transformation correctly before assembling them into larger systems.

\subsection{Bundle Inspired Data Containers}
\begin{table}[h!]
  \centering
\begin{tabular}{|lcl|}
  \hline \\
   fruit &  calories &  juice \\
  \hline\\
    apple &        95 &   True \\ 
   orange &        67 &   True \\ 
  lemon &        17 &  False \\ 
      lime &        20 &  False \\
  \hline
\end{tabular}
\label{tab:discussion:data}
\end{table}
We construct a toy dataset with a discrete $\dbase$ of 4 points and a fiber space of $\dfiber=\{apple,\, orange,\, lemon\}\times \mathbb{Z}^{+} \times \{\texttt{True}, \texttt{False}\}$. We thinly wrap \autoref{tab:discussion:data} in an object so that the common data interface function is that $\tau = \ttexttt{DataContainerObject.query}$. 
\begin{minted}{python}
  class FruitFrameWrapper:
    def query(self, data_bounds, sampling_rate):
      # local sections are a list of
      # {field: local_batch_of_values}
      return local_sections
\end{minted}

This interface provides a uniform way of accessing subsets of the data, which are local sections. The motivation for a common data interface is that it would allow the artist to talk to different common python data containers, such as numpy\cite{harris2020array}, pandas\cite{jeff_reback_2020_3715232}, xarray \cite{hoyer2017xarray}, and networkx\cite{HagbergExploringNetwork2008}. Currently, data stored in these containers must be unpacked and converted into arrays and matrices in ways that either destroy or recreate the structure encoded in the container. For example a pandas data frame must be unpacked into its columns before it is sent into most artists and continuity is implicit in the columns being the same length rather than a tracked base space $\dbase$. Because it is more efficient to work with the data in column order, we often treat the data as a collection of single fiber bundles. This is equivalent to the total bundle, as shown in \autoref{eq:artist:operator}.

\subsection{Component Encoders}
To encode the values in the dataset, we enforce equivariance by writing $\vchannel$ encoders that match the structure of the fields in the dataset. For example, the fruit column is a nominal measurement scale. Therefore we implement a position encoder that respects permutation $\dfunch$ transformations. The most simple form of this $\vchannel$ is a python dictionary that returns an integer position, because Matplotlib's internal parameter space expects a numerical position type. 
\begin{minted}{python}
  def position_encoder(val):
    return {'apple': 0, 'orange': 2, 'lemon': 4, 'lime': 6}[val]
\end{minted}
As mentioned in \autoref{eq:construction:nu:fabrication}, the encoders can be composed up. For example, the compositor $\vchannel$ may need the position to be converted to screen coordinates. Here the screen coordinate $\vchannel$ is a method of a Matplotlib axes object; a Matplotlib axes is akin to a container artist that holds all information about the sub artists plotted within it. 
\begin{minted}{python}
def composite_x_transform(ax, nu):
    return lambda x: ax.transData.transform(
            (position_encoder(x), 0))[0]
\end{minted}
This encoder returns a function that is \texttt{transData.transform} $\vchannel_{transData}$ composed with the position encoder $\vchannel_{position}$ and takes as input a record to be encoded. As with the position encoder, the transData encoder respects permutation transforms because it returns reals; therefore the composite encoder respects permutation transforms. In this model, developers implement $\vchannel$ encoders that are explicit about which $\dfunc_{\vtotal}$ they support. Writing semantically correct encoders is also the responsibility of the developer and is not addressed in the model. For example \mintinline{python}{fruit_encoder = lamda x: {'apple': green, 'orange':'yellow', 'lemon':'red', 'lime':'orange'}} is a valid color encoding with respect to permutation, but none of those colors are intuitive to the data. It is therefore left to the user, or domain specific library developer, to choose $\vchannel$ encoders that are appropriate for their data.

\subsection{Graphic Compositors}
After converting each record into an intermediate visual component $\vsection$, the set of visual records is passed into $\vmarkc$. Here the $\vmarkc$ includes one last encoder, as illustrated in \autoref{eq:construction:q:fabrication}, that assembles the independent visual components into a rectangle. This $\vchannel$ is inside the $\vmarkc$ to hide that library preferred format from the user. It is called \texttt{qhat} to indicate that this is the $\vartist^{\dbase}$ path in \autoref{eq:construction:artist:path}.  This means that the parameters are constructed in data space $\dbase$ and this function returns a pushed forward $\vindexpush\gsection$.

@@

\subsection{Integrating Components Into Existing Library}
The $\vchannel$ and $\vmark$ are wrapped in a container object that stores the $\vartist = \vmark \circ \vchannel$ composition and a method for computing the $\vsectionc$.
\begin{minted}{python}
  class Bar: 
    def compose_with_nu(self, pfield, ffield, 
          nu, nu_inv:):
        # returns a new copy of the Bar artist 
        # with the additional nu that converts 
        # from a data (F) field value to a 
        # visual (P) field value
        return new
